---
title: "Albemarle Equity Project"
author: "Charlotte McClintock"
date: "Spring 2019"
output: 
  html_document:
    toc: true
    theme: yeti
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE)
library(tidyverse)
```

## About The Project
***

The aim of this project is to explore demographic trends related to equity in Albemarle County to inform work on an equity profile of the county. 

### Initial Questions: 

  * Who lives in the county, and how is this changing?
  * Is the population growing?
  * Which groups are driving growth?
  * How diverse is the population?
  * How does the racial/ethnic composition vary by age?


## Data Sources
***
  * American Community Survey 5-Year (2013-2017), retrieved from [Social Explorer](https://www.socialexplorer.com/explore-maps) for county level age, sex, and race demographics, as well as median income, poverty rate. 
  * Small Area Income and Poverty Estimates, retrieved from [the Census Bureau](https://www.census.gov/data-tools/demo/saipe/saipe.html?s_appName=saipe&map_yearSelector=2017&map_geoSelector=aa_c)

# Demographics
***

Albemarle County has a population of 105,105 with 40,015 households. 

Average household size of 2.4 persons. 


Median Household Income of $72,265


### Race
```{r}
race <- read_csv("acs17-race.csv")
race$percent <-as.numeric(sub("%", "", race$percent))
ggplot(race, aes(x=reorder(race, percent), y=percent, fill=race)) +  
  stat_summary(fun.y="sum", geom="bar") + coord_flip() + guides(fill=F) + 
  geom_text(aes(y=percent, label=percent), hjust=-0.25) + ylim(0, 90)

```

### Age Demographics Relative to US
```{r, fig.width=11}
age <- read_csv("acs17-age.csv")
us.age <- read_csv("acs17-usage.csv") %>% select(-X4)
age$geo <- "Albemarle"
us.age$geo <- "United States"

age <- full_join(us.age, age, by=names(age))

age <- cbind(age, c(1:12))
age$percent <-as.numeric(sub("%", "", age$percent))
ggplot(age, aes(x=reorder(age, `c(1:12)`), y=percent, fill=geo)) +  
  stat_summary(fun.y="sum", geom="bar", position="dodge") + 
  geom_text(aes(y=percent, label=percent), vjust=-0.25, position = position_dodge(width = 1)) + 
  ylim(0, 16) + 
  theme(axis.text.x = element_text(angle = 90), legend.position = "bottom")

```


```{r}
ed <- read_csv("acs17-education.csv")
ed <- ed[-1,]
ed$percent <- as.numeric(sub("%", "", ed$percent))
ed$order <- c(1:7)

ggplot(ed, aes(x=reorder(ed_attain, -order), y=percent, fill=ed_attain)) + 
  stat_summary(fun.y="sum", geom="bar") + 
  coord_flip() + guides(fill=FALSE) + 
  geom_text(aes(y=percent, label=percent), hjust=-0.25)

```



## Economic Indicators 

### Median Household Income in Albemarle Relative to Virginia, America

```{r}
saipe.HI <- read_csv("saipe-medianHI.csv") %>% 
  select(Year, `State / County Name`, `Median Household Income in Dollars`)
names(saipe.HI) <- c("year", "geo", "medianHI")
saipe.HI <- mutate(saipe.HI, 
                    geo=as.factor(geo),
                    geo=fct_relevel(geo, 
                                    "Albemarle County (VA)",
                                    "Virginia",
                                    "United States"))

saipe.HI$medianHI <- as.numeric(str_replace_all(saipe.HI$medianHI, "[^[:alnum:]]", ""))

ggplot(saipe.HI, aes(x=year, y=medianHI, colour=geo)) + 
  geom_point(size=.5) + geom_line() + xlim(1989,2022) +
  annotate("text", x=2020, y=c(60336, 71518, 77591), label=c("$60,336", "$71,518", "$77,591"))
```


### Albemarle Median Household Income by Race

```{r, fig.width=10}
hi.race <- read_csv("acs17-hincomerace.csv")
hi.race <- filter(hi.race, !race=="Median Household Income (In 2017 Inflation Adjusted Dollars):"& !is.na(medianHI))
hi.race$medianHI <- str_replace_all(hi.race$medianHI, "[^[:alnum:]]", "")
hi.race$medianHI <- as.numeric(hi.race$medianHI)
ggplot(hi.race, aes(x=race, y=medianHI, fill=race)) + 
  stat_summary(fun.y="sum", geom="bar") + coord_flip() + 
  geom_hline(yintercept=72265, linetype=3) + 
  annotate("text", x=4, y=74000, label="Overall Median Household Income", angle=270) + 
  geom_text(aes(y=medianHI, label=medianHI), hjust=-0.25) + ylim(0, 85000) + guides(fill=F) 

```

### Poverty in Albemarle Relative to Virginia, America
```{r}
saipe.pov <- read_csv("saipe-poverty.csv") %>% 
  select(Year, `State / County Name`, `All Ages in Poverty Percent`)
names(saipe.pov) <- c("year", "geo", "povrate")
saipe.pov <- mutate(saipe.pov, 
                    geo=as.factor(geo),
                    geo=fct_relevel(geo, 
                                    "United States",
                                    "Virginia",
                                    "Albemarle County (VA)"))
ggplot(saipe.pov, aes(x=year, y=povrate, colour=geo)) + 
  geom_point(size=.5) + geom_line() + ylim(0,20) + 
  scale_x_continuous(breaks=c(1990, 1995, 2000, 2005, 2010, 2015, 2020)) + 
  annotate("text", x=2018.5, y=c(13.4, 10.7, 7.9), label=c("13.4%", "10.7%", "7.9%"))
```


### Albemarle Poverty Rate by Race

```{r, fig.width=11}
pov.race <- read_csv("acs17-povertyrace.csv") %>% select(race, income, n, perc)
pov.race <- pov.race[1:18,]

pov.race <- separate(pov.race, perc, sep="%", into="perc", extra="drop")
pov.race$perc <- as.numeric(pov.race$perc)

ggplot(pov.race, aes(x=race, y=perc, fill=income)) + 
  stat_summary(fun.y="sum", geom="bar", position="dodge") + 
  coord_flip() + theme(legend.position = "bottom") + scale_y_continuous(breaks=c(10,20,30,40,50,60,70,80,90,100))
```



