---
title: "Albemarle Equity Project"
author: "Charlotte McClintock"
date: "Spring 2019"
output: 
  html_document:
    toc: true
    theme: yeti
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE)
library(tidyverse)
```

## About The Project
***

The aim of this project is to explore demographic trends related to equity in Albemarle County to inform work on an equity profile of the county. 

### Initial Questions: 

  * Who lives in the county, and how is this changing?
  * Is the population growing?
  * Which groups are driving growth?
  * How diverse is the population?
  * How does the racial/ethnic composition vary by age?


## Data Sources
***
  * American Community Survey 5-Year (2013-2017), retrieved from [Social Explorer](https://www.socialexplorer.com/explore-maps) for county level age, sex, and race demographics, as well as median income, poverty rate. 
  * Small Area Income and Poverty Estimates, retrieved from [the Census Bureau](https://www.census.gov/data-tools/demo/saipe/saipe.html?s_appName=saipe&map_yearSelector=2017&map_geoSelector=aa_c)

# Demographics
***

Albemarle County has a population of 105,105 with 40,015 households and an average household size of 2.4 persons. 


### Race
```{r, fig.width=10, fig.height=5}
race <- read_csv("acs17-race.csv")
race$percent <-as.numeric(sub("%", "", race$percent))
ggplot(race, aes(x=reorder(race, percent), y=percent, fill=race)) +  
  stat_summary(fun.y="sum", geom="bar") + coord_flip() + guides(fill=F) + 
  geom_text(aes(y=percent, label=percent), hjust=-0.25) + ylim(0, 90) +
  labs(title="Race Demographics of Albemarle County (2017)",
       x="Percent of All Residents",
       y="Race", 
       caption="Data from the American Community Survey 5-Year (2013-2017)")

```

### Age Demographics Relative to US
```{r, fig.width=11}
age <- read_csv("acs17-age.csv")
us.age <- read_csv("acs17-usage.csv") %>% select(-X4)
age$geo <- "Albemarle"
us.age$geo <- "United States"

age <- full_join(us.age, age, by=names(age))

age <- cbind(age, c(1:12))
age$percent <-as.numeric(sub("%", "", age$percent))
ggplot(age, aes(x=reorder(age, `c(1:12)`), y=percent, fill=geo)) +  
  stat_summary(fun.y="sum", geom="bar", position="dodge") + 
  geom_text(aes(y=percent, label=percent), vjust=-0.25, position = position_dodge(width = 1)) + 
  ylim(0, 16) + 
  theme(axis.text.x = element_text(angle = 90), legend.position = "bottom")+
  labs(title="Age Demographics of Albemarle County vs. US Totals (2017)",
       x="Percent of All Residents",
       y="Age", 
       caption="Data from the American Community Survey 5-Year (2013-2017)",
       fill="Geography")

```

### Educational Attainment
```{r}
ed <- read_csv("acs17-education.csv")
ed <- ed[-1,]
ed$percent <- as.numeric(sub("%", "", ed$percent))
ed$order <- c(1:7)

ggplot(ed, aes(x=reorder(ed_attain, -order), y=percent, fill=ed_attain)) + 
  stat_summary(fun.y="sum", geom="bar") + 
  coord_flip() + guides(fill=FALSE) + 
  geom_text(aes(y=percent, label=percent), hjust=-0.25) +
  labs(title="Educational Attainment Demographics of Albemarle County (2017)",
       x="Percent of All Residents",
       y="Educational Attainment", 
       caption="Data from the American Community Survey 5-Year (2013-2017)")

```

### Household Demographics

```{r, fig.width=11, fig.height=4}
house <- read_csv("acs17-household.csv")
house <- separate(house, percent, sep="%", into="percent", extra="drop")

house <- mutate(house, 
                household_type=as.factor(household_type),
                household_type=fct_recode(household_type, 
   "Female Householder, No Spouse"="Female Householder, No Husband Present",
   "Male Householder, No Spouse"="Male Householder, No Wife Present"))
ggplot(subset(house, 
              !household_type=="Other Family:"&!household_type=="Family Households:"&!household_type=="Nonfamily Households:"), 
       aes(x=household_type, y=as.numeric(percent), fill=household_type)) + 
  stat_summary(fun.y = "sum", geom = "bar") + 
  theme(axis.text.x = element_text(angle = 90), legend.position = "bottom") + 
  coord_flip() + guides(fill=F) + 
  labs(title="Household Demographics for Albemarle County (2017)",
       y="Percent of All Households",
       x="Household Type", 
       caption="Data from the American Community Survey 5-Year (2013-2017)") + 
  geom_text(aes(label=percent, y=as.numeric(percent), hjust=-.25))
```


## Economic Indicators 

### Median Household Income in Albemarle Relative to Virginia, America

```{r, fig.width=8}
saipe.HI <- read_csv("saipe-medianHI.csv") %>% 
  select(Year, `State / County Name`, `Median Household Income in Dollars`)
names(saipe.HI) <- c("year", "geo", "medianHI")
saipe.HI <- mutate(saipe.HI, 
                    geo=as.factor(geo),
                    geo=fct_relevel(geo, 
                                    "Albemarle County (VA)",
                                    "Virginia",
                                    "United States"))

saipe.HI$medianHI <- as.numeric(str_replace_all(saipe.HI$medianHI, "[^[:alnum:]]", ""))

ggplot(saipe.HI, aes(x=year, y=medianHI, colour=geo)) + 
  geom_point(size=.5) + geom_line() + xlim(1989,2022) +
  annotate("text", x=2020, y=c(60336, 71518, 77591), label=c("$60,336", "$71,518", "$77,591")) +
  labs(title="Albemarle County, Virginia, and US Median Household Income (2017)",
       subtitle = "Albemarle has a consistently higher median income relative to US, VA",
       x="Year",
       y="Median Household Income", 
       caption="Data from the Census Small Area Income and Poverty Estimates",
       color="Geography")
```


### Albemarle Median Household Income by Race

```{r, fig.width=10}
hi.race <- read_csv("acs17-hincomerace.csv")
hi.race <- filter(hi.race, !race=="Median Household Income (In 2017 Inflation Adjusted Dollars):"& !is.na(medianHI))
hi.race$medianHI <- str_replace_all(hi.race$medianHI, "[^[:alnum:]]", "")
hi.race$medianHI <- as.numeric(hi.race$medianHI)
ggplot(hi.race, aes(x=race, y=medianHI, fill=race)) + 
  stat_summary(fun.y="sum", geom="bar") + coord_flip() + 
  geom_hline(yintercept=72265, linetype=3) + 
  annotate("text", x=4, y=74000, label="Overall Median Household Income", angle=270) + 
  geom_text(aes(y=medianHI, label=medianHI), hjust=-0.25) + ylim(0, 85000) + guides(fill=F) +
  labs(title="Income Demographics of Albemarle County by Race (2017)",
       x="Median Household Income",
       y="Race", 
       caption="Data from the American Community Survey 5-Year (2013-2017)")

```

### Poverty in Albemarle Relative to Virginia, America
```{r, fig.width=8}
saipe.pov <- read_csv("saipe-poverty.csv") %>% 
  select(Year, `State / County Name`, `All Ages in Poverty Percent`)
names(saipe.pov) <- c("year", "geo", "povrate")
saipe.pov <- mutate(saipe.pov, 
                    geo=as.factor(geo),
                    geo=fct_relevel(geo, 
                                    "United States",
                                    "Virginia",
                                    "Albemarle County (VA)"))
ggplot(saipe.pov, aes(x=year, y=povrate, colour=geo)) + 
  geom_point(size=.5) + geom_line() + ylim(0,20) + 
  scale_x_continuous(breaks=c(1990, 1995, 2000, 2005, 2010, 2015, 2020)) + 
  annotate("text", x=2018.5, y=c(13.4, 10.7, 7.9), label=c("13.4%", "10.7%", "7.9%"))+
  labs(title="Albemarle County, Virginia, and US Median Income (2017)",
       subtitle="Albemarle has a consistently lower poverty rate than VA, US",
       x="Year",
       y="Poverty Rate", 
       caption="Data from the Census Small Area Income and Poverty Estimates",
       color="Geography")
```


### Albemarle Poverty Rate by Race

```{r, fig.width=11, fig.height=6}
pov.race <- read_csv("acs17-povertyrace.csv") %>% select(race, income, n, perc)
pov.race <- pov.race[1:18,]

pov.race <- separate(pov.race, perc, sep="%", into="perc", extra="drop")
pov.race$perc <- as.numeric(pov.race$perc)

ggplot(filter(pov.race, income=="Income in Below Poverty Level"), aes(x=race, y=perc, fill=race)) + 
  stat_summary(fun.y="sum", geom="bar", position="dodge") + 
  coord_flip() + guides(fill=F) + scale_y_continuous(breaks=c(10,20,30,40,50,60,70,80,90,100)) + 
  geom_hline(yintercept = 7.9, linetype=2) + 
  scale_y_continuous(breaks=c(5,10,15,20,25)) +
  annotate("text", x=4.5, y=8.4, label="Overall Poverty Rate", angle=270) +
  labs(title="Poverty Rate by Race, Albemarle County (2017)",
       y="Poverty Rate",
       x="Race", 
       caption="Data from the American Community Survey 5-Year (2013-2017)")+ 
  geom_text(aes(y=perc, label=perc), hjust=-0.25) + ylim(0, 26)
```


### Income Distribution

```{r fig.width=9}
earn <- read_csv("acs17-earnings.csv", skip=1)
earn <- earn[, c(3, 40:147)]

earn <- gather(earn, 2:109, key="group", value="num")

earn <- separate(earn, group, sep=";", into=c("group", "type", "about"))
earn <- separate(earn, about, sep="EARNINGS - ", into=c("worker_type", "income_cat"))

earn <- mutate(earn, 
               income_cat=as.factor(income_cat),
               income_cat=fct_relevel(income_cat,
                                "$1 to $9,999 or loss",
                                "$10,000 to $14,999",
                                "$15,000 to $24,999", 
                                "$25,000 to $34,999",
                                "$35,000 to $49,999",
                                "$50,000 to $64,999", 
                                "$65,000 to $74,999", 
                                "$75,000 to $99,999", 
                                "$100,000 or more"))

ggplot(filter(earn, group=="Percent"&type==" Estimate"), aes(x=income_cat, y=num)) +
  stat_summary(fun.y="sum", geom="bar", fill="dark green") + 
  labs(title="Income Demographics for Albemarle County (2017)",
       y="Percent of All Earners",
       x="Income Category", 
       caption="Data from the American Community Survey 5-Year (2013-2017)")
```

### Income Distribution by Gender

```{r, fig.width=9}
ggplot(filter(earn, (group=="Percent Male"|group=="Percent Female")&type==" Estimate"), aes(x=income_cat, y=num, fill=group)) +
  stat_summary(fun.y="sum", geom="bar", position = "dodge") + 
  labs(title="Income Demographics by Gender for Albemarle County (2017)",
       y="Percent of All Earners",
       x="Income Category", 
       caption="Data from the American Community Survey 5-Year (2013-2017)")
```

